---
title: "Coffee Data Analysis"
author: "Laura Wiltshire"
date: "`r Sys.Date()`"
output: html_document
---

# Introduction
This analysis involves cleaning and processing a raw coffee dataset, 
imputing missing values, and performing initial exploratory data analysis 
(EDA) using R.
---

# Setup and Libraries
```{r setup, include=FALSE}
# Install and Load necessary libraries
# Define a function to install and load required packages
install_and_load <- function(packages) {
  for (pkg in packages) {
    if (!require(pkg, character.only = TRUE)) {
      install.packages(pkg, dependencies = TRUE)
      library(pkg, character.only = TRUE)
    }
  }
}

# List of required packages
required_packages <- c("tidyverse", "VIM", "e1071", "knitr", "rmarkdown")

# Install and load the required packages
install_and_load(required_packages)

```

# Load the Coffee Dataset
```{r Load the Dataset}
# Load the raw dataset
coffee_data <- read_csv("coffee-data-raw.csv")

# Extract structure of the dataset into a neat table
data_summary <- tibble(
  Variable = names(coffee_data),
  Type = sapply(coffee_data, class),
  Observations = sapply(
    coffee_data, 
    function(x) if (is.atomic(x)) length(x) else NA
  ),
  Missing = sapply(coffee_data, function(x) sum(is.na(x)))
)

# Display the table
kable(
  data_summary,
  caption = "Summary of the raw coffee dataset",
  col.names = c(
    "Variable", "Type", "Number of Observations", "Missing Values"
  ),
  format = "html"
)
```

# Impute Missing Values for 'roast'
```{r Impute missing roast values using k-means clustering}
set.seed(123)

coffee_data_imputed <- coffee_data %>%
  mutate(roast_imp = is.na(roast)) %>%
  VIM::kNN(variable = "roast", k = 5, imp_var = TRUE)

```

# Compare Original vs Imputed Values for Roast
```{r Compare original and imputed values of roast}
# Extract original and imputed roast values
original_values <- coffee_data_imputed %>% 
  filter(roast_imp == FALSE) %>% 
  pull(roast)

imputed_values <- coffee_data_imputed %>% 
  filter(roast_imp == TRUE) %>% 
  pull(roast)

# Plot density comparison of original and imputed values
ggplot(coffee_data_imputed, aes(x = roast, fill = roast_imp)) +
  geom_density(alpha = 0.6) + 
  scale_fill_manual(
    values = c("FALSE" = "#69b3a2", "TRUE" = "#404080"),
    labels = c("Original Values", "Imputed Values")
  ) +
  labs(
    x = "Roast Value",
    y = "Density",
    title = "Density Comparison of Original vs. Imputed Values",
    fill = "Value Type"
  ) +
  theme_minimal() +
  theme(
    legend.position = "top",
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    plot.title = element_text(size = 14, face = "bold"),
    panel.grid.minor = element_blank()
  )
```

# Generate Histogram of Coffee Ratings
```{r Histogram of Coffee Ratings}
ggplot(coffee_data_imputed, aes(x = rating)) +
  geom_histogram(binwidth = 2, fill = "brown", color = "black") +
  theme_minimal() +
  labs(
    x = "Rating Score",
    y = "Frequency"
  ) +
  theme(
    panel.grid = element_blank(),
    axis.title = element_text(size = 12), 
    axis.text = element_text(size = 10)
  )
```

# Generate Descriptive Statistics for Coffee Ratings
```{r Summary Statistics}
rating_summary <- data.frame(
  Statistic = c(
    "Minimum", "Maximum", "Mean", "Median", "Mode", 
    "Range", "Variance", "Standard Deviation", "IQR", 
    "Skewness", "Kurtosis"
  ),
  Value = c(
    min(coffee_data$rating, na.rm = TRUE),
    max(coffee_data$rating, na.rm = TRUE),
    mean(coffee_data$rating, na.rm = TRUE),
    median(coffee_data$rating, na.rm = TRUE),
    mode(coffee_data$rating), 
    paste(range(coffee_data$rating, na.rm = TRUE), collapse = "-"),
    var(coffee_data$rating, na.rm = TRUE),
    sd(coffee_data$rating, na.rm = TRUE),
    IQR(coffee_data$rating, na.rm = TRUE),
    skewness(coffee_data$rating, na.rm = TRUE),
    kurtosis(coffee_data$rating, na.rm = TRUE)
  )
)

kable(
  rating_summary,
  caption = "Summary Statistics for Coffee Ratings",
  col.names = c("Statistic", "Value"),
  format = "html"
)

```

# Save the Tidied Version of Coffee Data as .CSV
```{r Save the Final Version of Coffee Data}
rmarkdown::pandoc_available()

write_csv(coffee_data_imputed, "coffee-data-tidy.csv")

cat("The final dataset has been saved as 'coffee-data-tidy.csv'.\n")

rmarkdown::render("data-munge-markdown.Rmd")

```